Transferability of clinical prediction models for early trauma care

### 2019-10-13 ManuscriptMarkdownIJE.Rmd
Add files via upload

---
title: "Transferability of clinical prediction models for early trauma care"
author: 
- Martin Henriksson
- Dell D Saulnier
- Johanna Berg
- Martin Gerdin Wärnberg
date: "September 28, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyle.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

  *Background:* Trauma results in approximately four million deaths annually. Clinical
  prediction models are often used for triage in early trauma care. Previous research has
  indicated that transferring these models between different settings might adversely affect 
  model performance.

@@

  *Conclusions:* In all samples, model transfer was associated with some degree of uncertainty. 
  Models developed in samples with high numbers of patients performed best after transfer, and the
  model developed in the low-volume sample performed worst after transfer.

  **Keywords:** Trauma, Clinical prediction model, Transferability, Mistriage, Undertriage, Overtriage 
---

@@

## Setting
In Sweden, trauma is the most common cause of death before 44 years of age, and 
10% of major trauma patients suffer some form of disability. 
The societal effects of trauma are substantial, with the loss
of days at work comparable to those lost from cardiovascular and malignant diseases
combined [@RN23].

Trauma care in Sweden is usually clearly defined in major
metropolitan areas: trauma patients are transported directly to a predesignated
hospital with specific trauma competency. In more rural parts of the
country, trauma patients are instead transported to the nearest hospital for
stabilization. Once stabilized, they may then be transported to a hospital with
more advanced trauma care capacity [@RN34]. Most hospitals in Sweden use some
system to categorize the level of trauma, usually
based on patient vital signs and the mechanism of injury, as reported by the EMS or as
registered in the emergency department [@RN90].

Today, 52 of the 55 Swedish hospitals record trauma cases in SweTrau. 
Currently, the registry includes 55 000 cases, and its use is encouraged for 
both academic research and local quality improvement initiatives [@RN28].

@@

### Study outcome
ISS > 15 was used as the gold standard to define trauma severity as major
trauma, and hence, patients with ISS ≤ 15 were considered to have minor trauma
[@RN36]. Overtriage was defined as the event when a clinical prediction model
classified a patient with ISS ≤ 15 as having major trauma, and undertriage was defined as the event
when a clinical prediction model classified a patient with ISS > 15 as having minor
trauma. The overtriage rate was defined as the number of overtriaged patients
divided by the total number of patients. The undertriage rate was defined as the number of
undertriaged patients divided by the total number of patients. The mistriage rate was defined as the
sum of the over- and undertriage rates.

## Data sources and measurements
Model predictors, outcome, participant characteristics and study outcome as
outlined above were all obtained from SweTrau. The method of measurement of the
model predictors is not specified in the registry entries. In Swedish emergency rooms,
patient parameters are usually obtained by a registered nurse or assistant nurse
and are assumed to be accurate.  Whether the patient is dead or alive 30 days
after trauma is manually entered into the registry by each respective
hospital. The NISS and ISS are calculated by hospital personnel
based on the patient’s injuries.

## Bias
Data analysis was conducted according to a prearranged analysis plan. The
analysis plan and statistical analysis code were finalised using simulated
data. These efforts were made to avoid confirmation bias. The analysis plan was
reviewed by an experienced statistician and programmer prior to implementation. 
Neither the outcome nor variables were blinded during analysis, which 
made a structured approach important to ensure objectivity.

@@

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates. Varying degrees of uncertainty were observed when comparing 
the transferred performance to the receiving sample validation performance.

The most notable effect on model performance following model transfer was observed
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increased mistriage rate of `r substr(format(highest.comp.mistriage), 1, 4)`.
Mainly contributing to this was
an increase in overtriage.  For this transfer, the 95% UI ranged from 
`r UI.top.bot[1]` to `r UI.top.bot[2]`, indicating a high level of uncertainty. 
In contrast, the transfer of the high-volume model to the low-volume sample 
reduced mistriage, primarily due to reduced overtriage.

Low-volume centres do by definition receive fewer trauma patients, and therefore 
less data is available to establish accurate clinical prediction models. It is 
reasonable to assume this is the reason for the marked increase in mistriage 
when transferring the low-volume model to the high-volume sample. This also 
explains why the low-volume model and the single-centre model showed the highest 
levels of mistriage during model validation (i.e. when applied to the sample in 
which the model was created). Further confirming this, the transfer of the 
high-volume model to the low-volume sample reduced mistriage.

In a clinical context, this suggests that clinical prediction models for early 
trauma care should be developed in settings with a high number of trauma patients.
These models could then be transferred to settings with less trauma patients, which 
should then result in a decreased mistriage rate in those settings. In 2018, Granström 
et al. found that using a criteria-based triage system (i.e. a clinical prediction model) 
for trauma triage was effective in reducing overtriage without increasing mortality [@RN44]. 
Therefore, improving clinical prediction models to reduce mistriage is clinically beneficial, 
potentially saving lives by minimizing undertriage, and likely saving resources by 
minimizing overtriage. Possibly, with registries as SweTrau, the analysis of model 
transfer could be performed continuously to optimize the models currently in use and 
advise which model transfers might be detrimental (or potentially beneficial).

The transfer of clinical prediction models in trauma care has not previously
been studied extensively. In 2016, Gerdin et al. found that model transfer
adversely affected model performance in terms of calibration but that this could
be improved by updating the model [@RN27]. This study did not examine model
performance in terms of calibration, nor were the effects of updating the
model explored.

External validation - the evaluation of a clinical prediction model in a
setting in which it was not originally developed - has been studied more
extensively. Studies using simulated clinical prediction models based on
different predictors showed a decline in model performance when externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62] showed that the models
significantly underestimated the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest indiscriminate model transfers could lead to detrimental clinical
consequences, but also that clinically beneficial model transfers could be
identified.

@@

This study used GCS, RR and SBP as predictors for 30-day
mortality. Additional predictors such as age and mechanism
of injury or newer predictors such as the shock index might increase model
performance. Furthermore, we used 30-day mortality as an outcome, recognizing
that late mortality, time in hospital, functional impairment, and morbidity are
also important outcomes in the context of trauma care.

The predetermined analysis plan was to be followed in a step-by-step fashion. 
Programming challenges and late-time code inclusions required a flexible approach 
to this analysis plan, sometimes with revisions of previous steps. This is not 
optimal, and efforts were made to not deviate in any major way. Using GitHub, 
all code was made public for inspection and to ensure reproducibility [@RN59].

## Conclusion
Model transfer resulted in varying levels of uncertainty in in terms of its effect
on mistriage. In general, samples with a high number of patients had the lowest mistriage 
both during validation and after transfer to the other sample in the data set. The opposite 
was also true, and the transfer of the `r colnames(highest.comp.mistriage)` 
model led to an increased mistriage of `r highest.comp.mistriage`, noting that this observation 
was associated with considerable uncertainty. Recognizing the limitations of this study, 
the authors believe that further studies are warranted due to the potential economic 
costs and patient consequences of poor model transfers.

>>>>

---
title: "Transferability of clinical prediction models for early trauma care"
author:

date: "October 13, 2019"
documentclass: "article"
output:
  word_document:
    reference_docx: transferstyleIJE.docx
header-includes:
   - \usepackage{hyperref}
bibliography: library.bib
csl: bmcemerg.csl
---
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

  *Background:* Trauma results in approximately four million deaths annually. 
  Clinical prediction models are often used for triage in early trauma care. 
  This study aimed to assess how transfers of clinical prediction models for 
  early trauma care between different care contexts within a single health 
  system affected mistriage rates.

@@

  *Conclusions:* Models developed in samples with high numbers of patients generally 
  performed best after transfer, and the model developed in the low-volume sample 
  performed worst after transfer.

  **Keywords:** Trauma, Clinical prediction model, Transferability, Mistriage, Undertriage, Overtriage

  **Key Messages**

  * This study aimed to assess how transfers of clinical prediction models for early trauma care 
  between different care contexts within a single health system affected mistriage rates.

  * Three data sets, each with two samples,  were created representing plausible clinical 
  care contexts using data from the Swedish national trauma registry, SweTrau.

  *	Clinical prediction models were developed in each data set using logistic regression, 
  and then transferred within the data sets to simulate clinical prediction model transfer.

  * Models developed in samples with high-volumes of trauma patients generally performed 
  better, both in validation and after transfer.

  *	This study suggests that clinical prediction models for early trauma care developed 
  in settings with a high volume of trauma patients can be successfully transferred to 
  settings with lower volumes of trauma patients.

---

@@

## Setting
Today, 52 of the 55 Swedish hospitals record trauma cases in SweTrau. 
Currently, the registry includes 55 000 cases, and its use is encouraged for 
both academic research and local quality improvement initiatives [@RN28].

@@

### Study outcome
ISS > 15 was used as the gold standard to define trauma severity as major
trauma, and hence, patients with ISS â¤ 15 were considered to have minor trauma
[@RN36]. Overtriage was defined as the event when a clinical prediction model
classified a patient with ISS â¤ 15 as having major trauma, and undertriage was defined as the event
when a clinical prediction model classified a patient with ISS > 15 as having minor
trauma. The overtriage rate was defined as the number of overtriaged patients
divided by the total number of patients. The undertriage rate was defined as the number of
undertriaged patients divided by the total number of patients. The mistriage rate was defined as the
sum of the over- and undertriage rates.

## Data sources and measurements
Model predictors, outcome, participant characteristics and study outcome as
outlined above were all obtained from SweTrau. The method of measurement of the
model predictors is not specified in the registry entries. In Swedish emergency rooms,
patient parameters are usually obtained by a registered nurse or assistant nurse
and are assumed to be accurate.  Whether the patient is dead or alive 30 days
after trauma is manually entered into the registry by each respective
hospital. The NISS and ISS are calculated by hospital personnel
based on the patients injuries.

## Bias
Data analysis was conducted according to a prearranged analysis plan. The 
analysis plan was reviewed by an experienced statistician and programmer 
prior to implementation. Neither the outcome nor variables were blinded 
during analysis, which made a structured approach important to ensure objectivity.

@@

# Discussion
This study aimed to assess how transfers of clinical prediction models for early
trauma care between different contexts within a single health system affect
mistriage rates.

The most notable effect on model performance following model transfer was observed
after transferring the `r colnames(highest.comp.mistriage)` model. This transfer
resulted in an increased mistriage rate of `r substr(format(highest.comp.mistriage), 1, 4)`.
Mainly contributing to this was an increase in overtriage.
In contrast, the transfer of the high-volume model to the low-volume sample 
reduced mistriage, primarily due to reduced overtriage.

Low-volume centres do by definition receive fewer trauma patients, and therefore 
less data is available to establish accurate clinical prediction models. It is 
reasonable to assume this is the reason for the marked increase in mistriage 
when transferring the low-volume model to the high-volume sample. This also 
explains why the low-volume model and the single-centre model showed the highest 
levels of mistriage during model validation. Further confirming this, the transfer of the 
high-volume model to the low-volume sample reduced mistriage.

In a clinical context, this suggests that clinical prediction models for early 
trauma care should be developed in settings with a high number of trauma patients.
These models could then be transferred to settings with less trauma patients, which 
should then result in a decreased mistriage rate in those settings. In 2018, Granström 
et al. found that using a criteria-based triage system (i.e. a clinical prediction model) 
for trauma triage was effective in reducing overtriage without increasing mortality [@RN44]. 
Therefore, improving clinical prediction models to reduce mistriage is clinically beneficial, 
potentially saving lives by minimizing undertriage, and likely saving resources by 
minimizing overtriage.

The transfer of clinical prediction models in trauma care has not previously
been studied extensively. In 2016, Gerdin et al. found that model transfer
adversely affected model performance in terms of calibration but that this could
be improved by updating the model [@RN27].

External validation has been studied more extensively. Studies using simulated 
clinical prediction models based on
different predictors showed a decline in model performance when externally
validated [@RN60; @RN61]. A study from 2018 examining the external validity of
prediction models for coronary artery disease (CAD) [@RN62] showed that the models
significantly underestimated the probability of CAD. The combined results of
this study, the studies on model transfer and the studies on external validation
would suggest indiscriminate model transfers could lead to detrimental clinical
consequences, but also that clinically beneficial model transfers could be
identified.

@@

## Conclusion
Samples with a high number of patients had the lowest mistriage 
both during validation and after transfer to the other sample in the data set. The opposite 
was also true, and the transfer of the `r colnames(highest.comp.mistriage)` 
model led to a significantly increased mistriage, noting that this observation 
was associated with considerable uncertainty. Recognizing the limitations of this study, 
the authors believe that further studies are warranted due to the potential economic 
costs and patient consequences of poor model transfers.
